--- 
title: "GCAM Mapping Tools Examples"
author: "Catherine Ledna and Robert Link"
date: "18 December, 2106"
output:  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GCAM Mapping Tools Examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
--- 

# GCAM Mapping Tools Examples

## Introduction
This vignette explains how to use the GCAM Mapping Tools package to display GCAM data in map form.  Using this package has a couple of advangages over rolling your own maps.  First, it's easy.  By the time you've worked through these examples, you will be able to make maps of GCAM data with just a few simple commands.  Second, we have defined some default projections and extents for common use cases.  This will ensure that the maps you make look professional and conform the the GCAM house style.

```{r checknamespace, echo=FALSE}
if(!requireNamespace('rgcam',quietly=TRUE)) {
    stop('The rgcam package is required to build this vignette')
}
```

## Setup
To get started with the GCAM map tools, attach the `gcammaptools` package.  For now, you will also need to make the arcane-looking call to `gpclibPermit()` as shown below.
```{r setup}
library('gcammaptools')
## The function call below shouldn't be necessary, provided that we attach rgeos
## prior to attaching maptools.  We need to figure out how to make sure that
## happens in the library version.
maptools::gpclibPermit()                # TODO:  This function is deprecated.  Figure out how to replace.
```

## Loading GCAM data

You will need to load two types of data:  the map data (analogous to shape files in GIS tools) and the GCAM data.  The map data can be either loaded from the package (coming soon), or you can load your own.  Here is an example of how to do the latter.  The first line reads the map data, and the second turns it into a data frame indexed by region identifier.  This "map data frame" is what you will pass to the plotting functions.  Most of the map data frames you will need are provided as datasets in this package, but here is how you create one from a geojson file:
```{r load.map}
mapfile <- system.file('extdata/rgn32','GCAM_32_wo_Taiwan_clean.geojson', package = 'gcammaptools')
map_32_wo_Taiwan<-rgdal::readOGR(mapfile, "OGRGeoJSON")
map_32_wo_Taiwan.fort<-ggplot2::fortify(map_32_wo_Taiwan, region="GCAM_ID")
```
This gives you the same structure you would have gotten with `data(map.rgn32)`.  Generally you should use one of the map datasets included in the package whenever possible because they correspond to the geographical units reported by GCAM.  The datasets available are `map.rgn32`, `map.rgn14`, `map.basin235`, and `map.chn`.  They correspond to the 32-region and 14-region GCAM region maps, the 235 global water basin map, and the 32 regions plus China subregions map.

The best way to import GCAM results is to use the `rgcam` package, which will create a project file with your data in it.  You can load the project data with `rgcam::loadProject()` and retrieve tables for individual queries using `rgcam::getQuery()`.  However, if you aren't using `rgcam`, you can start with any data frame that has a `region` column and one or more data columns.

After loading the query you want to display, you must add the region identifiers used in the map data to the data frame using the `addRegionID()` function.  Then you can join the data and the map using `merge()`.

```{r load_data}
library('rgcam')
### Load the example scenario data.  We wrap it in suppressMessages() to hide an unhelpful notice
### about the file name of the sample data having changed.
suppressMessages({
    prj <- loadProject(system.file('sample-gcam-data','gcam-sample-project.dat', package='gcammaptools'))
})
listScenarios(prj)
listQueries(prj, 'Reference')
```

The sample data has two scenarios and and a dozen queries for each scenario.  For demonstration purposes we will load the CO<sub>2</sub> emissions table.  

```{r get_query}
co2 <- getQuery(prj, 'CO2 emissions by region', 'Reference')
co2 <- addRegionID(co2, lookupfile=rgn32, drops=rgn32)
map_co2<-merge(map_32_wo_Taiwan.fort, co2, by="id")
```
At the end of all this, `map_co2` is a map data frame that has primary energy consumption by region.  Each model year is a separate column.  This structure can be passed to `plot_GCAM` to plot maps.

## Sample Maps generated with plot_GCAM and GCAM32 shapefile

### Example 1: Eckert III World Projection, Colored by Region

This example just plots the map data frame with the GCAM region id.  Thus, you get each region colored according to a discrete color palette.
```{r mp1, fig.width=6, fig.height=3}
plot_GCAM(map_32_wo_Taiwan.fort, col = 'id', proj = eck3, colorfcn=qualPalette)
```

### Example 2: Robinson World Projection, Colored by Regional CO<sub>2</sub> Emissions
In this example we plot the map data frame of CO<sub>2</sub> that we created above.  We select the column for model year 2050.  We also use a different projection than we used above.
```{r mp2, fig.width=6, fig.height=4}
plot_GCAM(map_co2, col = "X2050", colors = c("white", "blue"),
          title="Robinson World", qtitle="CO2 Emissions, 2050", legend=T)
```

### Example 3: Winkel-Tripel Projection, Default Color and Style
This is a plot of the region boundaries in the Winkel-Tripel projection.  Since we didn't specify a column to plot, all of the regions are filled in grey.
```{r mp3, fig.width=6, fig.height=3}
plot_GCAM(map_32_wo_Taiwan.fort, proj=wintri, title="Winkel-Tripel World")
```

### Example 4: U.S. Projection (Albers Equal Area)
This map is specialized to the continental USA.  The `na_aea` and `EXTENT_USA` symbols are defined for convenience, but you can use any valid proj4 string (see `proj4::project` for how these strings are constructed) for the projection.  The extent should be the bounding box of the plot area in the form `c(lon.min, lon.max, lat.min, lat.max)`.
```{r mp4, fig.width=4, fig.height=4, fig.align='center'}
plot_GCAM(map_32_wo_Taiwan.fort, proj=na_aea,  extent=EXTENT_USA, title="USA Albers Equal-Area")
```

### Example 5: China Projection (Albers Equal Area)
A map of China.  Although the projection is once again the Albers equal area projection, we have to have a different projection string because the string includes some information about the parallels the projection is based on.
```{r mp5, fig.width=4, fig.height=4, fig.align='center'}
plot_GCAM(map_32_wo_Taiwan.fort, proj=ch_aea,extent=EXTENT_CHINA, title="China Albers Equal-Area")
```

### Example 6: Africa Projection (Orthographic)
For superregions with a long north-south extent, the orthographic projection gives the best result.  We use a different back-end function to compute the projection for this case, so you should pass the symbol `ortho` to the `proj` argument.  You will also have to supply the `orientation` argument, for which we have defined several convenient values.  This one gives you a map of Africa.
```{r mp6, fig.width=4, fig.height=4, fig.align='center'}
plot_GCAM(map_32_wo_Taiwan.fort, extent=EXTENT_AFRICA,
          orientation=ORIENTATION_AFRICA, proj=ortho,title= "Africa Orthographic")
```

### Example 7: Latin America Projection (Orthographic)
Orthographic projection of the Latin America superregion.
```{r mp7, fig.width=4, fig.height=4, fig.align='center'}
plot_GCAM(map_32_wo_Taiwan.fort, extent=EXTENT_LA,
        orientation= ORIENTATION_LA, proj=ortho, title="Latin America Orthographic")
```

### Example 8: Global Water basins
Here's an example using the global water basin map.
```{r mp8, fig.width=6, fig.height=3}
data(map.basin235)
plot_GCAM(map.basin235, col='id', colorfcn=qualPalette, proj=eck3)
```

