--- 
title: "GCAM Mapping Tools Examples"
author: "Caleb Braun, Catherine Ledna, and Robert Link"
date: "13 February, 2018"
output:  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GCAM Mapping Tools Examples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
--- 

# GCAM Mapping Tools Examples

## Introduction
This vignette explains how to use the GCAM Mapping Tools package to display GCAM data in map form.  Using this package has a couple of advantages over rolling your own maps.  First, it's easy.  By the time you've worked through these examples, you will be able to make maps of GCAM data with just a few simple commands.  Second, we have defined some default projections and extents for common use cases.  This will ensure that the maps you make look professional and conform to the GCAM house style.

```{r checknamespace, echo=FALSE}
if(!requireNamespace('rgcam', quietly=TRUE)) {
    stop('The rgcam package is required to build this vignette')
}
```

## Setup
To get started with the GCAM map tools, attach the `gcammaptools` package.
```{r setup}
library('gcammaptools')
```

## Loading GCAM data
You will need to load your own GCAM data to work with.  The best way to import GCAM results is to use the `rgcam` package, which will create a project file with your data in it.  You can then load the project data with `rgcam::loadProject()` and retrieve tables for individual queries using `rgcam::getQuery()`.  However, if you aren't using `rgcam`, you can start with any data frame that has a `region` column and one or more data columns.

```{r load_data, warning=FALSE}
library('rgcam')
### Load the example scenario data.
prj <- rgcam::loadProject(system.file('sample-gcam-data','gcam-longform-sample.dat',                                              package='gcammaptools'))
rgcam::listScenarios(prj)
rgcam::listQueries(prj, 'Reference')
```

The sample data has one scenario with 19 queries.  For demonstration purposes we will load the CO<sub>2</sub> emissions table. After loading the query you want to display, you must add the region identifiers used in the map data to the data frame using the `add_region_ID()` function.

```{r get_query}
co2 <- rgcam::getQuery(prj, 'CO2 emissions by region', 'Reference')
co2 <- dplyr::filter(co2, year==2050)
co2 <- gcammaptools::add_region_ID(co2, lookupfile=rgn32, drops=rgn32)

### Show the first observations
head(co2)
```
The data is in long form, so using `dplyr::filter()`  helps to get just the observations for the year you want to plot.  At the end of this, `co2` is a data frame that has CO<sub>2</sub> emissions by region for the year 2050.  This structure can now be passed to `plot_GCAM()` to plot maps, as shown in the examples below.

## Loading map data
When using `plot_GCAM()` for your GCAM data, you also need to provide the map data that it should be associated with.  You should use one of the map datasets included in the package whenever possible, because they correspond to the geographical units reported by GCAM.  The datasets available are `map.rgn32`, `map.rgn14`, `map.basin235`, `map.usa`, and `map.chn`.  They correspond to the 32-region and 14-region GCAM region maps, the 235 global water basin map, and the 32-region map plus states/provinces for the USA and China. Additionally, the non-provincial datasets each have a simplified version such as `map.rgn32.simple`, which is helpful for faster plotting and less cluttered maps. 
```{r plot.default, fig.width=3.48, fig.show='hold'}
gcammaptools::plot_GCAM(map.rgn32, title="Full 32-Region map")
gcammaptools::plot_GCAM(map.rgn32.simple, title="Simplified 32-Region map")
```

It is possible, however, for you to load your own map data as well.  The map data can be loaded by `gcammaptools` given the file path.  Your map data can be in any of the following formats: sf objects, spatial data frames, ESRI Shapefiles, or GeoJSON files.  You can either pass the file path to `plot_GCAM()` or load the data yourself first.
```{r load.map, fig.width=4, fig.align='center'}
mapdata <- system.file("extdata/rgn32", "reg32_spart.shp", package = "gcammaptools")
mapdata
gcammaptools::plot_GCAM(mapdata)
```

## Sample maps

### Example 1: Eckert III World Projection, Colored by Region

This example just plots the map data frame with the GCAM region name.  Applying a scale with `gcam32_colors`, the default GCAM colors, you get each region colored according to a discrete color palette.  For older data that uses 14-region GCAM, you can use the palette `gcam14_colors`.
```{r mp1, fig.width=6, fig.height=3, fig.align='center'}
gcammaptools::plot_GCAM(map.rgn32.simple, col = 'region_name', proj = eck3) +
    ggplot2::scale_fill_manual(values = gcam32_colors, na.value=gray(0.75))
```

### Example 2: Robinson World Projection, Colored by Regional CO<sub>2</sub> Emissions
In this example we plot the CO<sub>2</sub> data frame that we created above.  We select the column `value`, which is the name of the column that contains the data.  It is also necessary to specify how to join the this data frame to the map data, which is why we created the column `id` with the function `add_region_id()` above.
```{r mp2, fig.width=6, fig.height=4, fig.align='center'}
gcammaptools::plot_GCAM(map.rgn32.simple, col='value', proj=robin, title="Robinson World", legend=T,
          gcam_df=co2) + 
    ggplot2::scale_fill_gradientn(colors = c("white", "red"),
                                  na.value = gray(0.75), name="CO2 Emissions (MTC)")
```

### Example 3: U.S. Projection (Albers Equal Area)
This map is specialized to the continental USA.  The `na_aea` and `EXTENT_USA` symbols are defined for convenience, but you can use any valid proj4 string (see `proj4::project` for how these strings are constructed) for the projection.  The extent should be the bounding box of the plot area in the form `c(lon.min, lon.max, lat.min, lat.max)`.
```{r mp3, fig.width=4, fig.height=4, fig.align='center'}
gcammaptools::plot_GCAM(map.rgn32, col='region_id', proj=na_aea, extent=EXTENT_USA, 
          title="USA Albers Equal-Area")
```

### Example 4: Africa Projection (Orthographic)
For superregions with a long north-south extent, the orthographic projection gives the best result. We have predefined one for Africa, but you can find others defined by proj4 strings from http://spatialreference.org as shown in the following example. Although the extent parameter is the best way to specify the view you want, you can fine-tune the final plot by adjusting the zoom.
```{r mp4, fig.width=4, fig.height=4, fig.align='center'}
gcammaptools::plot_GCAM(map.rgn32, col='region_name', proj=af_ortho, extent=EXTENT_AFRICA,
          title="Africa Orthographic") +
    ggplot2::scale_fill_manual(values = gcam32_colors)
```

### Example 5: Latin America Projection (Orthographic)
Orthographic projection of the Latin America superregion. Notice that projection strings can also be defined by specifying an EPSG, ESRI, or SR-ORG projection code.
```{r mp5, fig.width=4, fig.height=4, fig.align='center'}
gcammaptools::plot_GCAM(map.rgn32, col='region_name', proj=7567, proj_type='SR-ORG',
          extent=EXTENT_LA, title="Latin America Orthographic") +
    ggplot2::scale_fill_manual(values = gcam32_colors)
```

### Example 6: China Projection (Albers Equal Area)
A map of China.  Although the projection is once again the Albers equal area projection, we have to have a different projection string because the string includes some information about the parallels the projection is based on.
```{r mp6, fig.width=4, fig.height=4, fig.align='center'}
gcammaptools::plot_GCAM(map.rgn32, proj=ch_aea, extent=EXTENT_CHINA, title="China Albers Equal-Area")
```


### Example 7: Global Water Basins
A map of the 235 global water basins.  Because `col` is set to `basin_name`, each basin name is treated as a separate category, giving the colorful map below.
```{r mp7, fig.width=6, fig.height=3, fig.align='center'}
gcammaptools::plot_GCAM(map.basin235, col='basin_name', proj=eck3)
```


### Example 8: Gridded Data
The `plot_GCAM_grid()` function tiles gridded data over a base map.
```{r mp8, fig.width=4, fig.height=6, fig.align='center'}
co2grid <- rgcam::getQuery(prj, 'Cooling Degree Days', 'Reference')
gcammaptools::plot_GCAM_grid(co2grid, col='value', proj=robin, extent=EXTENT_LA, legend=T) + 
    ggplot2::scale_fill_gradientn(colors=c("white", "red"), 
                                  guide=ggplot2::guide_colorbar(title="Cooling Degree Days", 
                                                                title.position="top"))

```

### Example 9: Faceting
Using `ggplot::facet_wrap()` is often a good way to compare the results from several different scenarios. To do this with `plot_GCAM()`, it is important to make sure that your dataset contains the facet variable for each region you want to plot. By default, `add_region_ID()` puts in `NA` values for missing regions. The `disaggregate` parameter allows you to specify that the missing regions with no data should be included in each facet.
```{r mp9, fig.width=7, fig.height=3, fig.align='center'}
waterdata <- read.csv(system.file('extdata','facet_example.csv', package='gcammaptools'),
                      stringsAsFactors = F)
waterdata <- gcammaptools::add_region_ID(waterdata, disaggregate = 'scenario')

gcammaptools::plot_GCAM(map.rgn32.simple, col='value', legend=T, gcam_df=waterdata,
          title="Water Scarcity in 2100") + 
    ggplot2::scale_fill_gradientn(colors=c("deepskyblue", "firebrick1"),
                                  na.value=gray(0.9),
                                  guide=ggplot2::guide_colorbar(title="km^3",
                                                                barwidth=12)) +
    ggplot2::facet_wrap(~scenario)

```

### Example 10: GCAM China
Data from GCAM China may contain extra regions that you want to drop as well as region name abbreviations that you want to translate into the full province name.
```{r mp10, fig.width=6, fig.height=4, fig.align='center'}
gcamchina <- read.csv(system.file('extdata', 'china_example.csv', package='gcammaptools'),
                      stringsAsFactors = F)
gcamchina <- gcammaptools::add_region_ID(gcamchina, lookupfile = chn, provincefile = chn, drops = chn)

# Highlight the provinces by filtering out values from other regions
gcamchina[!is.na(gcamchina$id) & gcamchina$id <= 32, 'value'] <- NA
gcammaptools::plot_GCAM(map.chn, col='value', gcam_df=gcamchina, proj=ch_aea,
          extent=EXTENT_CHINA)
```

### Example 11: GCAM USA
Support is also provided for GCAM USA data. The map expects states to be identified by their two letter abbreviations, which can be added on using `add_region_id()` with `lookupfile = usa`, as long as the state full names are present.
```{r mp11, fig.width=6, fig.height=4, fig.align='center'}
gcamusa <- read.csv(system.file('extdata', 'usa_example.csv', package='gcammaptools'),
                    stringsAsFactors = F)
gcammaptools::plot_GCAM(map.usa, col='value', gcam_df=gcamusa, gcam_key='region', proj=na_aea,
          extent=EXTENT_USA)
```
```{r}
## Example 12 Plot refined liquids production by region for the year 2050
prj <- loadProject(system.file('sample-gcam-data',
                                'gcam-longform-sample.dat',
                                package='gcammaptools'))
 ref_liquids <- rgcam::getQuery(prj, 'Refined liquids production by region', 'Reference')
 ref_liquids <- add_region_ID(ref_liquids, lookupfile=rgn32, drops=rgn32)
 ref_liquids <- dplyr::filter(ref_liquids, year==2050)
 plot_GCAM(map.rgn32.simple, col='value', proj=robin, title="Robinson World",
           legend=T, gcam_df=co2, gcam_key='id', mapdata_key="region_id") +
    ggplot2::scale_fill_gradientn(colors = c("white", "red"),
                                  na.value = gray(0.75),
                                  name="CO2 Emissions (MTC)")
```

